Precise Data Flow Analysis\\in the Presence of Correlated Method Calls

### 2015-03-08 sas2015/content/abstract.tex
minor tweaks

	We show how infeasible paths due to correlated calls can be eliminated for 
	\textit{Interprocedural Finite Distributive Subset} (IFDS) problems, a large class 
	of data-flow analysis problems with broad applications. Our approach is to
	transform an IFDS problem into an \textit{Interprocedural Distributive 
	Environment} (IDE) problem, in which edge functions are designed in a
	way that enables us to ignore data flow along infeasible paths. A solution
	to this IDE problem can be mapped back to the solution space of the original
	IFDS problem.
        We formalize the approach, prove it correct, and report on an implementation
        in the WALA analysis framework.
\end{abstract}

>>>>

	We show how infeasible paths due to correlated calls can be eliminated for 
	\textit{Interprocedural Finite Distributive Subset} (IFDS) problems, a large class 
	of data-flow analysis problems with broad applications. Our approach is to
	transform an IFDS problem into an \textit{Interprocedural Distributive 
	Environment} (IDE) problem, in which edge functions ``filter out'' data flow 
	along infeasible paths. A solution
	to this IDE problem can be mapped back to the solution space of the original
	IFDS problem.
        We formalize the approach, prove it correct, and report on an implementation
        in the WALA analysis framework.
\end{abstract}

### 2015-03-08 sas2015/content/background.tex
minor tweaks

\subsection{Terminology and Notation}
%We will start by introducing several concepts used by the IFDS and IDE analyses.
%\begin{mdelete}\mtodo{We already say (but less precisely) what a CFG is in the introduction.}
The \textit{control-flow graph} of a procedure is a directed graph whose nodes are instructions, and which contains an edge from $n_1$ to $n_2$ whenever
$n_2$ may execute immediately after $n_1$. A CFG has a distinguished \textit{start node} \textsf{start}$_p$ and \textit{end node} \textsf{end}$_p$.
Following the presentation of Reps~et~al.~\cite{reps1995precise,sagiv1996precise}, we follow every
call instruction with a no-op instruction, so that every \textit{call node} is immediately
followed by a \textit{return node} in the CFG.
%\end{mdelete}
%and edges represent transfer of control between the instructions during an execution of the program. 
%A control-flow graph has a unique start node, $\startmain$, which is the node corresponding to the program entrypoint.
%An \textit{intraprocedural} path is a path in a control-flow graph whose nodes are in the same procedure. By contrast, an \textit{interprocedural path} is one that contains nodes from different procedures. 
The \textit{control-flow supergraph} of a program contains the CFGs of all of the procedures as
subgraphs. In addition, for each call instruction $c$, the supergraph contains a \textit{call-to-start} edge to the start node of every procedure that
may be called from $c$, and an \textit{end-to-return} edge from the end node of the procedure back to the call instruction.
%\otodo{do we really need to extra "return nodes" that the IFDS/IDE papers add?}
%is a control-flow graph
%in which each procedure $p$ is augmented with an additional \textit{start node} \textsf{start}$_p$ and \textit{end node} \textsf{end}$_p$, and for each call $c_q$ to a procedure $q$, there is a \textit{call node} \textsf{call}$_{c_q}$ and subsequent \textit{return node} \textsf{return}$_{c_q}$.

@@

A path through the control-flow supergraph is \textit{valid} if it follows the usual stack-based calling 
discipline: every end-to-return edge on the path returns
to the site of the most recent call that has not yet been matched by a return. 
%\ftodo{Can we explain/define the notion of the ``most recent unmatched call'' more precisely?}
The set of all valid paths from the program entry point to
a node $n$ is denoted $\ivp(n)$.
%\otodo{Should we give a formal definition?}

@@

%\otodo{The definitions that Marianna had were for \emph{complete}
%lattices/semilattices. All of the lattices in the context of IFDS/IDE
%are/must be complete. I'll just say that here.}
%\ftodo{Would it make sense to first the define the notion of a lattice without referring to its completeness?
       %This could be followed by defining what it means for a (semi)lattice to be complete, and then we
       %could say that all lattices considered in this paper are complete, etc.}
%\ftodo{Can we define the symbols $\sqcup$ and $\sqcap$ for join and meet here?}
A \textit{lattice}%
\footnote{The definitions that we give here are of \textit{complete} lattices and semilattices.
Since all of the (semi)lattices discussed in this paper are required to be complete, we omit the
\textit{complete} qualifier.}
is a partially ordered set $(S,\sqsubseteq)$ in which every subset has a least upper bound, called \textit{join} or $\sqcup$, and a greatest lower bound, called \textit{meet} or $\sqcap$.
A \textit{meet semilattice} is a partially ordered set in which every subset need only have a greatest lower bound.
The symbols $\bot$ and $\top$ are used to denote the greatest lower bound of $S$ and of the empty set, respectively.

@@

\subsection{IFDS}\label{sec:bgifds}
The IFDS framework~\cite{reps1995precise} is a precise and efficient algorithm for data-flow analysis
%. IFDS was developed in 1995 by T.\,Reps, S.\,Horwitz, and M.\,Sagiv at the University of Wisconsin and
that has been used to solve a variety of data-flow analysis problems~\cite{bodden2013spl,naeem2008typestate,DBLP:conf/birthday/KreikerRRSWY13,tripp2009taj}.
%\mtodo{ The four references take away space in the References section. Remove some of them?  }
The IFDS framework is an instance of the \textit{functional approach} to data-flow analysis~\cite{pnueli1981two}
because it constructs summaries of the effects of called procedures.
\commentout{The IFDS analysis is a version of the classic \textit{functional approach} to data-flow analysis proposed by M.\,Sharir and A.\,Pnueli~\cite{pnueli1981two}.}
\commentout{In other data-flow algorithms not based on the functional approach, the result of the analysis at the entry of a procedure ``merges'' the incoming data obtained from all callers of the procedure. As a consequence, there is one global data-flow result computed at the end of the procedure. Context-sensitivity, however, allows an analysis to compute the data-flow result for a given procedure as a \textit{function} of the data-flow value at the start of the procedure. 
In other words, the analysis result for a procedure varies depending on where the procedure was called from. This significantly improves the precision of a data-flow analysis, which is why context-sensitivity is an important advantage of IFDS over classic data-flow algorithms.}
The IFDS framework is applicable to \textit{interprocedural} data-flow problems whose domain is \textit{subsets} of a \textit{finite} set $D$,
and whose data-flow functions are \textit{distributive}.
A function $f$ is distributive if 
$f(x_1\sqcap x_2)=f(x_1)\sqcap f(x_2)$.

@@

\begin{example}\label{ex:flowfun}
    Given $D=\{u,\,v,\,w\}$ and $f(S) = S\setminus\{v\} \cup \{u\}$,
    the representation relation 
  %The representation relation $R_f$ for a set of data-flow facts 
  %and data-flow function $f=\lambda S\,.\,S\setminus\{v\}\cup\{u\}$ looks as follows:
  $
    R_f=\{(\mathbf0,\,\mathbf0),\,(\mathbf0,\,u),\,(w,\,w)\}
$, which can be visualized with the following graph:
   \begin{center}
    \tikzstyle{n}=[fill=black,circle,inner sep=1.5pt]
\begin{tikzpicture}[>=stealth,on grid, auto]
    \node [n,label=$\mathbf{0}$] (zero){};
    \node [n,] (zero_) [below=\edge of zero] {};
    \node [n,label=$u$] (u) [right=\nd of zero] {};
    \node [n,] (u_) [below=\edge of u] {};
    \node [n,label=$v$] (v) [right=\nd of u] {};
    \node [n,] (v_) [below=\edge of v] {};
    \node [n,label=$w$] (w) [right=\nd of v] {};
    \node [n,] (w_) [below=\edge of w] {};
    \path[->](zero) edge (zero_);
    \path[->](zero) edge (u_);
    \path[->](w) edge (w_);
  \end{tikzpicture}
  \end{center}
\end{example}

@@

\input{figures/fig-cc_edge_fn_example}
%\mtodo{In Figure~\ref{fig:cc_edgefn_example}, should we say that $\psi$ corresponds to the function's return value?}
\begin{example}
  The exploded supergraph for Listing~\ref{list:ccexample} is shown in Figure~\ref{fig:cc_edgefn_example}.
  The labels on the edges will be explained in Section~\ref{sec:bgide}
  We can see that there is a realizable path, highlighted in bold,
  from the start node of the exploded graph to the variable \verb's' at~the~node~$\textsf{print(s)}$ in the \verb'B.bar' method. This means that at that node, \verb's' is considered secret.

@@

The composition of two environment transformers can be computed by combining their pointwise
representation graphs in the same way as IFDS representation relations, and 
computing the composition of the micro-functions appearing along each path in the resulting graph.

>>>>

\subsection{Terminology and Notation}
%We will start by introducing several concepts used by the IFDS and IDE analyses.
%\begin{mdelete}\mtodo{We already say (but less precisely) what a CFG is in the introduction.}
The \textit{control-flow graph} of a procedure is a directed graph whose nodes are instructions, which contains an edge from $n_1$ to $n_2$ whenever
$n_2$ may execute immediately after $n_1$. A CFG has a distinguished \textit{start node} \textsf{start}$_p$ and \textit{end node} \textsf{end}$_p$.
Following the presentation of Reps~et~al.~\cite{reps1995precise,sagiv1996precise}, we follow every
call instruction with a no-op instruction, so that every \textit{call node} is immediately
followed by a \textit{return node} in the CFG.
%\end{mdelete}
%and edges represent transfer of control between the instructions during an execution of the program. 
%A control-flow graph has a unique start node, $\startmain$, which is the node corresponding to the program entrypoint.
%An \textit{intraprocedural} path is a path in a control-flow graph whose nodes are in the same procedure. By contrast, an \textit{interprocedural path} is one that contains nodes from different procedures. 
The \textit{control-flow supergraph} of a program contains the CFGs of all of the procedures as
subgraphs. In addition, for each call instruction $c$, the supergraph contains a \textit{call-to-start} edge to the start node of every procedure that
may be called from $c$, and an \textit{end-to-return} edge from the end node of the procedure back to the call instruction.
%\otodo{do we really need to extra "return nodes" that the IFDS/IDE papers add?}
%is a control-flow graph
%in which each procedure $p$ is augmented with an additional \textit{start node} \textsf{start}$_p$ and \textit{end node} \textsf{end}$_p$, and for each call $c_q$ to a procedure $q$, there is a \textit{call node} \textsf{call}$_{c_q}$ and subsequent \textit{return node} \textsf{return}$_{c_q}$.

@@

A path in the control-flow supergraph is \textit{valid} if it follows the usual stack-based calling 
discipline: every end-to-return edge on the path returns
to the site of the most recent call that has not yet been matched by a return. 
%\ftodo{Can we explain/define the notion of the ``most recent unmatched call'' more precisely?}
The set of all valid paths from the program entry point to
a node $n$ is denoted $\ivp(n)$.
%\otodo{Should we give a formal definition?}

@@

%\otodo{The definitions that Marianna had were for \emph{complete}
%lattices/semilattices. All of the lattices in the context of IFDS/IDE
%are/must be complete. I'll just say that here.}
%\ftodo{Would it make sense to first the define the notion of a lattice without referring to its completeness?
       %This could be followed by defining what it means for a (semi)lattice to be complete, and then we
       %could say that all lattices considered in this paper are complete, etc.}
%\ftodo{Can we define the symbols $\sqcup$ and $\sqcap$ for join and meet here?}
A \textit{lattice}%
\footnote{The definitions that we give here are of \textit{complete} lattices and semilattices.
Since all of the (semi)lattices discussed in this paper are required to be complete, we omit the
\textit{complete} qualifier.}
is a partially ordered set $(S,\sqsubseteq)$ in which every subset has a least upper bound, called \textit{join} or $\sqcup$, and a greatest lower bound, called \textit{meet} or $\sqcap$.
A \textit{meet semilattice} is a partially ordered set in which every subset only has a greatest lower bound.
The symbols $\bot$ and $\top$ are used to denote the greatest lower bound of $S$ and of the empty set, respectively.

@@

\subsection{IFDS}\label{sec:bgifds}
The IFDS framework~\cite{reps1995precise} is a precise and efficient algorithm for data-flow analysis
%. IFDS was developed in 1995 by T.\,Reps, S.\,Horwitz, and M.\,Sagiv at the University of Wisconsin and
that has been used to solve a variety of data-flow analysis problems~\cite{bodden2013spl,naeem2008typestate,DBLP:conf/birthday/KreikerRRSWY13,tripp2009taj}.
%\mtodo{ The four references take away space in the References section. Remove some of them?  }
The IFDS framework is an instance of the \textit{functional approach} to data-flow analysis~\cite{pnueli1981two}
because it constructs summaries of the effects of called procedures.
\commentout{The IFDS analysis is a version of the classic \textit{functional approach} to data-flow analysis proposed by M.\,Sharir and A.\,Pnueli~\cite{pnueli1981two}.}
\commentout{In other data-flow algorithms not based on the functional approach, the result of the analysis at the entry of a procedure ``merges'' the incoming data obtained from all callers of the procedure. As a consequence, there is one global data-flow result computed at the end of the procedure. Context-sensitivity, however, allows an analysis to compute the data-flow result for a given procedure as a \textit{function} of the data-flow value at the start of the procedure. 
The analysis result for a procedure varies depending on where the procedure was called from. This significantly improves the precision of a data-flow analysis, which is why context-sensitivity is an important advantage of IFDS over classic data-flow algorithms.}
The IFDS framework is applicable to \textit{interprocedural} data-flow problems whose domain is \textit{subsets} of a \textit{finite} set $D$,
and whose data-flow functions are \textit{distributive}.
A function $f$ is distributive if 
$f(x_1\sqcap x_2)=f(x_1)\sqcap f(x_2)$.

@@

\begin{example}\label{ex:flowfun}
    Given $D=\{u,\,v,\,w\}$ and $f(S) = S\setminus\{v\} \cup \{u\}$,
    the representation relation 
  %The representation relation $R_f$ for a set of data-flow facts 
  %and data-flow function $f=\lambda S\,.\,S\setminus\{v\}\cup\{u\}$ looks as follows:
  $
    R_f=\{(\mathbf0,\,\mathbf0),\,(\mathbf0,\,u),\,(w,\,w)\}
$, which is depicted by the following graph:
   \begin{center}
    \tikzstyle{n}=[fill=black,circle,inner sep=1.5pt]
\begin{tikzpicture}[>=stealth,on grid, auto]
    \node [n,label=$\mathbf{0}$] (zero){};
    \node [n,] (zero_) [below=\edge of zero] {};
    \node [n,label=$u$] (u) [right=\nd of zero] {};
    \node [n,] (u_) [below=\edge of u] {};
    \node [n,label=$v$] (v) [right=\nd of u] {};
    \node [n,] (v_) [below=\edge of v] {};
    \node [n,label=$w$] (w) [right=\nd of v] {};
    \node [n,] (w_) [below=\edge of w] {};
    \path[->](zero) edge (zero_);
    \path[->](zero) edge (u_);
    \path[->](w) edge (w_);
  \end{tikzpicture}
  \end{center}
\end{example}

@@

\input{figures/fig-cc_edge_fn_example}
%\mtodo{In Figure~\ref{fig:cc_edgefn_example}, should we say that $\psi$ corresponds to the function's return value?}
\begin{example}
  The exploded supergraph for Listing~\ref{list:ccexample} is shown in Figure~\ref{fig:cc_edgefn_example}.
  The labels on the edges will be explained in Section~\ref{sec:bgide}
  We can see that there is a realizable path, highlighted in bold,
  from the start node of the exploded graph to the variable \verb's' at~the~node~$\textsf{print(s)}$ in the \verb'B.bar' method. This means that \verb's' is considered secret at that node.

@@

The composition of two environment transformers can be computed by combining their pointwise
representation graphs in the same fashion as IFDS representation relations, and 
computing the composition of the micro-functions appearing along each path in the resulting graph.

### 2015-03-08 sas2015/content/example.tex
minor tweaks

In this section, we illustrate our approach using a small motivating example that
shows how our technique can be used to improve the precision of taint analysis.
A taint analysis is a specialized data-flow analysis for computing how string values
may flow from ``sources'', which are typically statements that read untrusted input, 
to ``sinks'', which are typically security-sensitive
operations such as calls to 
a database. In previous research~\cite{DBLP:conf/issta/GuarnieriPTDTB11,DBLP:conf/pldi/ArztRFBBKTOM14}, 
taint analysis algorithms have been formulated as IFDS problems.

@@

Figure~\ref{list:ccexample} shows a small Java program that we will use to illustrate
how our technique can improve the precision of a simple taint analysis. The program
declares a class \code{A} with a subclass \code{B},  where \code{A} defines methods 
\code{foo()} and \code{bar()} that are overridden in \code{B}.  In order to keep the 
example small and self-contained, it is assumed  that secret values are created by 
an unspecified function \code{secret()}, which is called in method \code{A.foo()} on 
line~\ref{line:Afoo}. Furthermore, it is assumed that any write to the standard output 
stream via calls \code{System.out.println()} such as the one in method \code{B.bar()} 
on line~\ref{line:Bbar} is a security-sensitive operation. Depending on the number of 
arguments passed to the program, the \code{main()} method of the example program creates 
either an \code{A}-object or a \code{B}-object on line~\ref{line:createObject}. The 
program then calls \code{foo()} on this object on line~\ref{line:callfoo}, which is 
followed by a call to \code{bar()} on the same object on line~\ref{line:callbar}.

@@

Informally speaking, an IFDS-based data-flow analysis propagates data-flow facts along the edges
of a control flow supergraph such as the one in Figure~\ref{fig:examplesupergraph}. The
IFDS algorithm was carefully designed to avoid infeasible paths that may arise in the
presence of multiple calls to the same method. However, in this example all methods are
called in exactly one place, so IFDS is unable to eliminate data flow along any of the
paths shown in the figure. As a result, IFDS-based taint analysis algorithms such as
\cite{DBLP:conf/issta/GuarnieriPTDTB11,DBLP:conf/pldi/ArztRFBBKTOM14} would report 
that the secret value read on line~\ref{line:Afoo}
might flow to the print statement on line~\ref{line:Bbar}.

@@

The main contribution of this paper is an algorithm for transforming an IFDS problem 
into an IDE problem that expresses the feasibility of paths
in light of correlated calls.
Informally speaking, our approach associates with each edge in an
interprocedural CFG a function that records the types of 
variables that are used as the receiver of correlated method calls. Paths that 
are composed of edges in which the same receiver expression has different types
are infeasible, and the propagation of data-flow facts along such paths is
prevented. Applying our technique to an IFDS-based taint analysis would enable
the resulting IDE-based taint analysis to determine that no secret value can flow from
line~\ref{line:Afoo} to the print statement on line~\ref{line:Bbar}.

>>>>

In this section, we illustrate our approach using a small motivating example that
shows how our technique improves the precision of taint analysis.
A taint analysis is a specialized data-flow analysis for computing how string values
may flow from ``sources'', which are typically statements that read untrusted input, 
to ``sinks'', which are typically security-sensitive
operations such as calls to 
a database. In previous research~\cite{DBLP:conf/issta/GuarnieriPTDTB11,DBLP:conf/pldi/ArztRFBBKTOM14}, 
taint analysis algorithms have been formulated as IFDS problems.

@@

Figure~\ref{list:ccexample} shows a small Java program that we will use to illustrate
how our technique improves the precision of a simple taint analysis. The program
declares a class \code{A} with a subclass \code{B},  where \code{A} defines methods 
\code{foo()} and \code{bar()} that are overridden in \code{B}.  In order to keep the 
example small and self-contained, it is assumed  that secret values are created by 
an unspecified function \code{secret()}, which is called in method \code{A.foo()} on 
line~\ref{line:Afoo}. Furthermore, it is assumed that any write to the standard output 
stream via calls \code{System.out.println()} such as the one in method \code{B.bar()} 
on line~\ref{line:Bbar} is a security-sensitive operation. Depending on the number of 
arguments passed to the program, the \code{main()} method of the example program creates 
either an \code{A}-object or a \code{B}-object on line~\ref{line:createObject}. The 
program then calls \code{foo()} on this object on line~\ref{line:callfoo}, which is 
followed by a call to \code{bar()} on the same object on line~\ref{line:callbar}.

@@

An IFDS-based data-flow analysis propagates data-flow facts along the edges
of a control flow supergraph such as the one in Figure~\ref{fig:examplesupergraph}. The
IFDS algorithm was carefully designed to avoid infeasible paths that may arise in the
presence of multiple calls to the same method. However, in this example all methods are
called in exactly one place, so IFDS is unable to eliminate data flow along any of the
paths shown in the figure. As a result, IFDS-based taint analysis algorithms such as
\cite{DBLP:conf/issta/GuarnieriPTDTB11,DBLP:conf/pldi/ArztRFBBKTOM14} would report 
that the secret value read on line~\ref{line:Afoo}
might flow to the print statement on line~\ref{line:Bbar}.

@@

The main contribution of this paper is an algorithm for transforming an IFDS problem 
into an IDE problem that expresses the feasibility of paths
in light of correlated calls.
Our approach associates with each edge in an
interprocedural CFG a function that records the types of 
variables that are used as the receiver of correlated method calls. Paths that 
are composed of edges in which the same receiver expression has different types
are infeasible, and the propagation of data-flow facts along such paths is
prevented. Applying our technique to an IFDS-based taint analysis would enable
the resulting IDE-based taint analysis to determine that no secret value can flow from
line~\ref{line:Afoo} to the print statement on line~\ref{line:Bbar}.

### 2015-03-08 sas2015/content/intro.tex
minor tweaks

This paper presents an approach to data-flow analysis that avoids a type
of infeasible path that arises in object-oriented programs when two or more
methods are dynamically dispatched on the same receiver object.
%\mtodo{Should we define the term ``receiver''?}
In such cases, if the
method calls are polymorphic (i.e., if they dispatch to different method definitions 
depending on the type of the receiver expression at run time), then their dispatch 
behaviours will be correlated. A recent paper~\cite{DBLP:journals/scp/Tip15} identified
this problem but did not present a concrete solution or algorithm, and we are not aware 
of any existing data-flow analysis that is capable of avoiding infeasible paths that arise 
in the presence of correlated method calls.

>>>>

This paper presents an approach to data-flow analysis that avoids a type
of infeasible path that arises in object-oriented programs when two or more
methods are dynamically dispatched on the same receiver object.
%\mtodo{Should we define the term ``receiver''?}
In such cases, if the
method calls are polymorphic (i.e., if they dispatch to different method definitions 
depending on the type of the receiver expression at run time), then their dispatch 
behaviours will be correlated. A recent paper~\cite{DBLP:journals/scp/Tip15} identified
this problem but did not present a concrete solution or algorithm, and we are not aware 
of any existing data-flow analysis that is capable of avoiding infeasible paths that arise 
in the presence of correlated method calls (see Section~\ref{sec:Related} for a discussion
of related work).
